---
title: "US Economic Time Series Analysis"
author: "Divya Pariti - 23200831"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
```

**Introduction**

This report explores the US economic time series dataset available in the `ggplot2` package. The dataset provides several economic indicators over time, which are crucial for understanding trends in the US economy.

**Overview of the Dataset**

Here are the main components of the economics dataset:

1.  `date` : The date of the observation, recorded monthly.
2.  `pce` : Personal consumption expenditures (in billions of dollars). This measures the total amount of spending by individuals in the U.S. on goods and services.
3.  `pop` : Total population of the U.S. at the time of the observation.
4.  `psavert` : Personal saving rate, which is the percentage of people's income remaining each month after taxes and spending.
5.  `uempmed` : Median duration of unemployment, in weeks. This measures the median time that unemployed people have been looking for work.
6.  `unemploy` : Number of unemployed individuals, in thousands.

```{r load-data, echo=FALSE}
head(economics)
```

```{r}
# Display a summary of the 'economics' dataset.
# The summary includes statistics like minimum, maximum, median, mean, and quartiles for each variable.
summary(economics)
```

```{r}
# Compute correlation matrix
corr_mat = cor(economics[c("pce", "pop", "psavert", "uempmed", "unemploy")])

# Print correlation matrix
print(corr_mat)
```
**Personal Consumption Expenditures **

```{r}
# Boxplot of Personal Consumption Expenditures (PCE)
ggplot(economics, aes(x = 1, y = pce)) +
  geom_boxplot(fill = "burlywood") +
  labs(x = "", y = "PCE in billion dollars",
       title = "Boxplot of PCE over time") +
  theme_minimal()
```

**Economic Trends Over Time**

***Unemployment Rate***

The following plot shows the unemployment rate over time, which is a key indicator of economic health.

```{r unemployment-plot, echo=FALSE, fig.cap="Figure 1: Unemployment Rate Over Time"}
ggplot(economics, aes(x = date, y = unemploy / pop)) +
  geom_line(color = "blue") +
  labs(x = "Year", y = "Unemployment Rate (as a fraction of population)",
       title = "Unemployment Rate Over Time") +
  theme_minimal()
```

**Analysis:**

-   This plot illustrates the changes in the unemployment rate over time.

-   Understanding these trends is vital for economic planning and policy-making.

-   The graph begins with a relatively low unemployment rate, which slowly rises.

-   Before the 2000s, there is a period of relative stability with lower unemployment rates, suggesting a healthy economic phase.

-   After the peak in 2009, the unemployment rate shows a declining trend, the rate decreases steadily, suggesting economic improvement and job market stabilization.

**Conclusion:**

-   The `economics` dataset comprises monthly records of several key economic indicators such as unemployment rate, personal savings rate, and personal consumption expenditures.

-   The unemployment rate is a crucial indicator of economic health, reflecting underlying economic conditions, policy changes, and global economic impacts.

-   The fluctuations observed in the graph correspond with historical economic events, demonstrating the sensitivity of employment to economic cycles.

-   The overall downward trend in recent years suggests effective economic policies and recovery mechanisms.
